The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-14 kalman_walk_paper.tex
overleaf edits

@@

Above we show how system variation and drift can contribute to gene network incompatibility.
What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.

%% What is A and what determines its genetic variance.
The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
per change in concentration of $j$.
It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
In the context of transcription factor networks this may be affected 
not only by the binding strength of molecule $j$ to the promoter region of gene $i$
but also the effects of other transcription factors (e.g., cooperativity)
and local chromatin accessibility \citep{stefflova2013cooperativity}.
For this reason, 
the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
and single variants may affect many entries of $\Sys$ simultaneously.
% (Recall that although these are best modeled through nonlinear terms,
% by linearizing we essentially consider first-order effects.)
On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
in the presence of substantial molecular noise \citep{canalization,buffering},
suggesting that substantial variation in the micro-scale dynamics we consider here
may be necessary to produce relevant phenotypic effects downstream.
\plr{replace "micro-scale" with sthg else or discuss earlier}

%% empirical data on sigma
The amount and structure of this standing variation is established over long time scales
by many factors, including
mutation-selection balance, 
shifts in the phenotypic optimum,
and/or spatial variation in the optimum \citep{hansen1996translating}.
Quantitative genetics models of mutation-selection balance 
predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
but it is unclear how well these predictions match reality \citep{johnson_barton}
and how much they are expected to change over time \citep{arnold_changing_G}.
However, empirical work allows us to estimate at least the rough magnitude of variation.

Differences in $A_{ij}$ due to a sequence change are hard to measure,
but variation in both transcription factor binding site occupancy
and expression levels (e.g., cis-eQTL) have been measured in various systems.
However, variation in binding site occupancy may overestimate variation in $\Sys$,
since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless,
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citep{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $\Sys$
may be on the order of 1\% between individuals of a population --
doubtless varying substantially between species and between genes.

What matters is $\sigma$ relative to $\gamma$ but this is harder to know.

@@

% nonlinearity
What about nonlinearity?
As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}. As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations. Most notable is linearity. It is often stated that life is not linear. This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory. Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzansky-Muller incompatibilities.

>>>>

%%% CITE SOME OF THESE
%% empirical data on sigma
% The amount and structure of this standing variation is established over long time scales
% by many factors, including mutation-selection balance, 
% shifts in the phenotypic optimum, and/or spatial variation in the optimum \citep{hansen1996translating}.
% Quantitative genetics models of mutation-selection balance 
% predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
% but it is unclear how well these predictions match reality \citep{johnson_barton}
% and how much they are expected to change over time \citep{arnold_changing_G}.

@@

What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.
The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
per change in concentration of $j$.
It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
In the context of transcription factor networks this may be affected 
not only by the binding strength of molecule $j$ to the promoter region of gene $i$
but also the effects of other transcription factors (e.g., cooperativity)
and local chromatin accessibility \citep{stefflova2013cooperativity}.
For this reason, 
the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
and single variants may affect many entries of $\Sys$ simultaneously.
% (Recall that although these are best modeled through nonlinear terms,
% by linearizing we essentially consider first-order effects.)
% On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
% in the presence of substantial molecular noise \citep{canalization,buffering},
% suggesting that substantial variation in the micro-scale dynamics we consider here
% may be necessary to produce relevant phenotypic effects downstream.

Variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless,
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citep{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $\Sys$
may be on the order of at least a few percent between individuals of a population --
doubtless varying substantially between species and between genes.

The impact of regulatory variation ($\sigma$) above
depends only on its magnitude relative to selection ($\gamma$),
but it seems reasonable that these two quantities are of the same magnitude.

@@

% nonlinearity
What about nonlinearity?
As Richard Levins opined, models in population biology face a trade-off among precision, realism, and generality \citep{levins1966strategy}. As Levins expects, any tractable and general model, such as the present one under discussion, will have limitations. Most notable is linearity. It is often stated that life is not linear. This is often true, however, many of the ideas developed here should be generalizable to nonlinear cases (multi-linear systems, say). Further, we see this as a necessary first step in the direction of more life-like nonlinear evolutionary systems theory. Depending on an actual biological system's particularities, its (potential) non-linearity, may buffer or exacerbate effects elucidated in this paper, such as the acquisition of Dobzhansky-Muller incompatibilities.
